---
title: DATA 606 Final Project
author: "Josh Iden
---

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(knitr)
library(GGally)
library(infer)
```


### Part 1 - Introduction
This project focuses on the question "Is education level predictive of voter attitudes amongst non-voters?" using data collected for the FiveThirtyEight.com report ["Why Many Americans Don't Vote"](https://projects.fivethirtyeight.com/non-voters-poll-2020-election/)

### Part 2 - Data
The dataset contains survey data from 5,836 respondents, collected from polling done by Ipsos for FiveThirtyEight. The study is observational, relying on data taken from a random sample. Details about the data collection can be found [here](https://github.com/fivethirtyeight/data/blob/master/non-voters/README.md)

### Part 3 - Exploratory data analysis
First, let's take a look at the data.

```{r}
#loading packages
library(tidyverse)

# loading dataset
url = "https://raw.githubusercontent.com/fivethirtyeight/data/master/non-voters/nonvoters_data.csv"
df = read.csv(url)

dim(df)
```

The dataset contains 5,836 observations across 119 columns (questions). For the purpose of this analysis, we will subset the data down to the desired data: Education, and Questions 5, 16, and 20.

These questions are as follows:

**Q5: Does it matter who wins elections? 1 = Yes, 2 = No**  
**Q16: In general, how easy or difficult is it to vote in elections? 1 = Very Easy, 4 = Very Difficult**  
**Q20: Are you currently registered to vote? 1 = Yes, 2 = No**  


```{r}
cols = c("educ","Q5","Q16","Q20")
df.sub = df[cols]
kable(head(df.sub))
```

Let's take a look at the summary statistics

```{r}
summary(df.sub)
```
```{r}
sum(is.na(df.sub))
```

The first thing that pops out are the minimum values of -1, since the minimum response value is 1. Let's see how many of those there are.

```{r}
df.sub |> filter(Q5 == -1 | Q16 == -1 | Q20 == -1) |> count()
```
There are 82 rows with -1 entries.

```{r}
df.sub |> filter(Q5 == -1 | Q16 == -1 | Q20 == -1) |> head()
```
We can see that these negative values do not otherwise affect other responses. This is likely a data entry error. To rectify this, I will impute a value of 1 for every -1 value. 

```{r}
df.sub <- df.sub |>
  mutate(Q5 = abs(Q5),
            Q16 = abs(Q16),
            Q20 = abs(Q20))

summary(df.sub)
```
We can see this didn't significantly affect the summary statistics.

Let's take a look at the distribution of education.

```{r}
ggplot(df.sub, aes(x=educ)) +
  geom_bar() + 
  geom_text(aes(label = after_stat(count)), 
            stat="count", vjust=2,
            color="white") +
  labs(x = "education")
```

```{r}
kable(prop.table(table(df.sub$educ)), col.names=c("Level","Freq"))
```

We can see from the survey responses, college graduates accounted for just under 40% of non-voters amongst respondents, while those hadn't completed college accounted for just over 60% of respondents.

Taking a look at average answer by education level,

```{r}
df.sub |> 
  group_by(educ) |>
  summarize(Q5 = mean(Q5),
            Q16 = mean(Q16),
            Q20 = mean(Q20))
```

We can see that responders with an education level of "High School or less" has the highest response scores -- with higher scores representing more negative attitudes towards voting. Somewhat surprisingly, respondents with "some college" tend to view the difficulty involved with voting (Q16) more favorably than the other groups. 

Let's take a look at the distribution of responses by education level.

```{r}
df.sub |>
  pivot_longer(cols = c(2:4), names_to = "question", values_to = "rating") |>
  ggplot(aes(x=question, y=rating)) +
  facet_wrap(~ educ) +
  geom_boxplot()
```

We can see that because the values are so closely disbersed, the distributions are challenging to visualize. 

### Part 4 - Inference

First, let's convert the educational column to a dichotomous variable - Yes or No. 

```{r}
college <- df.sub |>
  mutate(college = ifelse(educ == "College", "yes", "no")) |>
  select(5, 2:4)

kable(head(college))
```
```{r}
kable(prop.table(table(college$college)), col.names=c("Level","Freq"))
```

Now let's estimate the population proportion.

```{r}
college |>
  specify(response = college, success = "yes") |>
  generate(reps = 1000, type = "bootstrap") |>
  calculate(stat = "prop") |>
  get_ci(level = 0.95)
```
This tells us that 95% of the samples of the population of non-voters have mean college degree proportions between 38.6% and 41.1%

Let's see how college education predicts the answers to the survey questions. To do this, we'll create dummy variables for college degree

```{r}
college <- college |>
  mutate(college_dummy = ifelse(college == "yes", 1, 0))

lm1 <- glm(college_dummy ~ Q5 + Q16 + Q20, data=college)
summary(lm1)
```
We can tell right off the bat that Q16, perceived difficulty in voting, is not statistically significant, looking first at its p-value, so we remove it from the model and rerun. 

```{r}
lm2 <- glm(college_dummy ~ Q5 + Q20, data=college)
summary(lm2)
```


### Part 5 - Conclusion


### References


### Appendix (optional)

Remove this section if you don't have an appendix
