---
title: "DATA 606 Data Project Proposal"
author: "Josh Iden"
output: pdf_document
---

### Data Preparation

```{r setup, echo=TRUE, results='hide', warning=FALSE, message=FALSE}
# load data
library(tidyverse)

url = "https://raw.githubusercontent.com/fivethirtyeight/data/master/non-voters/nonvoters_data.csv"
df = read.csv(url)

head(df)
```


### Research question 

**Is education level predictive of voter attitudes amongst non-voters?**

### Cases 

**The dataset contains survey data from 5,836 respondents**

### Data collection 

**The data was collected from polling done by Ipsos for FiveThirtyEight for the story [Why Many Americans Don't Vote](https://projects.fivethirtyeight.com/non-voters-poll-2020-election/). Details about the data collection can be found [here](https://github.com/fivethirtyeight/data/blob/master/non-voters/README.md)**

### Type of study 

**This study is observational, relying on data taken from a random sample survey.**

### Data Source 

**The data is courtest of [FiveThirtyEight](https://github.com/fivethirtyeight/data/tree/master/non-voters)**

### Dependent Variable

**The response variables are different attitudes about elections, ie, "Do elections matter?", "How easy or difficult is it to vote?", and current voter registration.**

### Independent Variable(s)
**The independent variable is level of education**

### Relevant summary statistics 

**Provide summary statistics for each the variables. Also include appropriate visualizations related to your research question (e.g. scatter plot, boxplots, etc). This step requires the use of R, hence a code chunk is provided below. Insert more code chunks as needed.**

```{r}
cols = c("educ","Q5","Q16","Q20")
df_sub = df[cols]
head(df_sub)
```

```{r}
ggplot(df_sub, aes(educ)) +
  geom_bar(stat="count")
```
Does it matter who wins elections? 1 = Yes, 2 = No

```{r}
Q5 = df_sub |>
  select(c(educ, Q5)) |>
  group_by(educ) |>
  summarize(mean = mean(Q5))

Q5
```
```{r}
ggplot(Q5, aes(x=educ, y=mean)) +
  geom_bar(stat="identity")
```
*We can see from a high level, proportionally, those with less education are more skeptical of the difference voting makes*

In general, how easy or difficult is it to vote in elections? 1= Very Easy, 4 = Very Difficult

```{r}
Q16 <- df_sub |>
  select(c(educ, Q16)) |>
  group_by(educ) |>
  summarize(mean = mean(Q16))

Q16
```
```{r}
ggplot(df_sub, aes(Q16, fill=educ)) +
  geom_bar(stat="count")
```
*Surprisingly, those with some college have more favorable opinions regarding the ease of voting than those with college degrees, although respondents with high school educations or less find voting harder*

Are you currently registered to vote? 1 = Yes, 2 = No

```{r}
Q20 <- df_sub |>
  select(c(educ, Q20)) |>
  group_by(educ) |>
  summarize(mean = mean(Q20))

Q20
```
```{r}
ggplot(df_sub, aes(x=Q20, fill=educ)) +
  geom_bar(stat="count", position="dodge")
```

*We can see that college educated respondents have the highest proportion of voter registartion, followed by those with some college. Respondents with high school educations or less have the lowest proportion of voter registration*

*We can also see from the data above that there is likely a data entry issue with a -1 appearing in place of 1, to be addressed in tidying*
